Mahotas: Open source software for scriptable computer vision

### 2012-09-01 paper.tex
TXT Readthrough fixes

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information and performing basic manipulations and
computations. In particular, unlike libraries written in the C~Language or in
Java~\citep{Marcel:2010:TMP:1873951.1874254}, Mahotas does not need to define a
data structure, but uses the numpy array structure. Many basic manipulation
functionality that would otherwise be part of a computer vision library are
handled by numpy, for example computing averages and other simple statistics,
handling multi-channel images, converting between types (integer and floating
point images are supported by mahotas, whenever it is meaningful). For the
user, this has the additional advantage that they do not need to learn yet
another set of functions.

@@

Version 1.0 of mahotas has been released recently and this is now a mature,
well-tested package (the first version was made available over 4~years ago).
\footnote{Note for reviewers: the version currently available implements all
functionality described in this manuscript. I will release it as version 1.0
when this manuscript is accepted to coincide with publication. Naturally, any
bugs that are found and reported in the meanwhile will be addressed.} Mahotas
runs and is used on different versions of Unix (including Linux, SunOS, and
FreeBSD), Mac OS X, and Windows.

@@

\begin{description}
\item[\textsc{surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[wavelet] Haar and Daubechies wavelets. Forward and inverse transforms.
\item[morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[filtering] Gaussian filtering and general convolutions.
\item[polygon operations] convex hull, polygon drawing.
\end{description}

@@

When a new image is to be returned, functions take an argument named \code{out}
where the output will be stored. This argument is often much more restricted in
type. In particular, it must often be a contiguous array.\footnote{Numpy
supports non-contiguous arrays, which are most often slices into other, larger,
contiguous arrays (e.g., given a $128 \times 128$ contiguous array, one can
build a $64 \times 128$ non-contiguous array by taking every other row).} Since
this is a performance feature (its purpose is to avoid extra memory
allocation), it is natural that the interface is less flexible (accessing a
contiguous array is much more efficient than a non-contiguous one).

@@

\begin{cplusplus}
PyObject* py_erode(PyObject* self, PyObject* args) {
    PyArrayObject* array;
    PyArrayObject* Bc;
    PyArrayObject* output;
    if (!PyArg_ParseTuple(args, "OOO", &array, &Bc, &output) ||
        !numpy::are_arrays(array, Bc, output) || !numpy::same_shape(array, output) ||
        !PyArray_EquivTypenums(PyArray_TYPE(array), PyArray_TYPE(Bc)) ||
        !PyArray_EquivTypenums(PyArray_TYPE(array), PyArray_TYPE(output)) ||
        PyArray_NDIM(array) != PyArray_NDIM(Bc)
    ) {
        PyErr_SetString(PyExc_RuntimeError, TypeErrorMsg);
        return NULL;
    }
    holdref r_o(output);

@@

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{SAFE\_SWITCH\_ON\_INTEGER\_TYPES()} macro, which call the right
specialisation of the template that does the actual work. In this example
\code{erode} implements (binary) erosion:

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that release the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done. This allows several
mahotas functions to run concurrently.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject} that
knows its data type and has iterators which resemble the \cpp{} standard
library. This makes the code type-safer. This is also a \textsc{raii} object in
terms of managing Python reference counts.
\item[\code{filter\_iterator}] This is taken from \code{scipy.ndimage} and it
is useful to iterate over an image and use a centered filter around each pixel
(it keeps track of all of the boundary conditions).
\end{description}

@@

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar (data not
shown).

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, all types are checked and an exception is generated which can
be caught by user code). This is particularly relevant for interactive use as
the user is often exploring and is willing to pay the cost of a few extra type
conversions to avoid a hard-crash.

@@

\textbf{Additional system requirements}\\
None.

@@

\textbf{Code repository}\\
\textit{Name}: Github\\
\textit{Identifier}: https://github.com/luispedro/mahotas\\
\textit{Licence}: MIT\\
\textit{Date published}: Since 2009

@@

\textbf{Funding}: I was supported in my work by the Funda\c c\~{a}o para a
Ci\^encia e Tecnologia (grants SFRH/BD/37535/2007 and
PTDC/SAU-GMG/115652/2008), by NIH grant GM078622 (to Robert F.\ Murphy), by a
grant from the Scaife Foundation, by the HHMI Interface initial, and by a grant
from the Siebel Scholars Foundation.

>>>>

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3 and up). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information and performing basic manipulations and
computations. In particular, unlike libraries written in the C~Language or in
Java~\citep{Marcel:2010:TMP:1873951.1874254}, Mahotas does not need to define a
new image data structure, but uses the numpy array structure. Many basic
manipulation functionality that would otherwise be part of a computer vision
library are handled by numpy, for example computing averages and other simple
statistics, handling multi-channel images, converting between types (integer
and floating point images are supported by mahotas). For the user, this has the
additional advantage that they do not need to learn yet another set of
functions.

@@

Version 1.0 of mahotas has been released recently and this is now a mature,
well-tested package (the first version was made available over 4~years ago).
\footnote{Note for reviewers: the version currently available (0.9.2)
implements all functionality described in this manuscript. I will release it as
version 1.0 when this manuscript is accepted to coincide with publication.
Naturally, any bugs that are found and reported in the meanwhile will be
addressed.} Mahotas runs and is used on different versions of Unix (including
Linux, SunOS, and FreeBSD), Mac OS X, and Windows.

@@

\begin{description}
\item[\textsc{surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[wavelet] Haar and Daubechies wavelets. Forward and inverse transforms.
\item[morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[filtering] Gaussian filtering, edge finding, and general convolutions.
\item[polygon operations] convex hull, polygon drawing.
\end{description}

@@

When a new image is to be returned, functions take an argument named \code{out}
where the output will be stored. This argument is often much more restricted in
type. In particular, it must be a contiguous array.\footnote{Numpy supports
non-contiguous arrays, which are most often slices into other, larger,
contiguous arrays (e.g., given a $128 \times 128$ contiguous array, one can
build a $64 \times 128$ non-contiguous array by taking every other row).} Since
this is a performance feature (its purpose is to avoid extra memory
allocation), it is natural that the interface is less flexible (accessing a
contiguous array is much more efficient than a non-contiguous one).

@@

\begin{cplusplus}
PyObject* py_erode(PyObject* self, PyObject* args) {
    PyArrayObject* array;
    PyArrayObject* Bc;
    PyArrayObject* output;
    if (!PyArg_ParseTuple(args, "OOO", &array, &Bc, &output) ||
        !numpy::are_arrays(array, Bc, output) || !numpy::same_shape(array, output) ||
        !numpy::equiv_typenums(array, Bc, output) ||
        PyArray_NDIM(array) != PyArray_NDIM(Bc)
    ) {
        PyErr_SetString(PyExc_RuntimeError, TypeErrorMsg);
        return NULL;
    }
    holdref r_o(output);

@@

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{SAFE\_SWITCH\_ON\_INTEGER\_TYPES()} macro, which call the right
specialisation of the template that does the actual work. In this example
\code{erode} implements erosion:

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that releases the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done. This allows several
mahotas functions to run concurrently.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject}, the raw
numpy data type, which has iterators which resemble the \cpp{} standard
library. It also handles type-casting internally, making the code type-safer.
This is also a \textsc{raii} object in terms of managing Python reference
counts.
\item[\code{filter\_iterator}] This is taken from \code{scipy.ndimage} and it
is useful to iterate over an image and use a centered filter around each pixel
(it keeps track of all of the boundary conditions).
\end{description}

@@

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image)}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar (data not
shown).

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, all types are checked and an exception is generated which can
be caught by user code). This is particularly relevant for interactive use as
the user is often exploring and is willing to pay the speed cost of a few extra
type conversions to avoid a hard-crash.

@@

\textbf{Additional system requirements}\\
None at runtime. Compilation from source requires a \cpp{} compiler.

@@

\textbf{Code repository}\\
\textit{Name}: Github\\
\textit{Identifier}: https://github.com/luispedro/mahotas\\
\textit{Licence}: MIT\\
\textit{Date published}: Since 2010

@@

\textbf{Funding}: I was supported in my work by the Funda\c c\~{a}o para a
Ci\^encia e Tecnologia (grants SFRH/BD/37535/2007 and
PTDC/SAU-GMG/115652/2008), by NIH grant GM078622 (to Robert F.\ Murphy), by a
grant from the Scaife Foundation, by the HHMI Interfaces Initiative, and by a
grant from the Siebel Scholars Foundation.
